Surrogate waveform model using Gaussian process regression for an aligned-spin effective one body model of binary neutron star inspirals

### 2018-10-23 paper/paper.tex
Paper : added a paragraph on quadrupole-monopole terms.

For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are those of the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins. \red{It includes spin-orbit interactions up to 3PN order and spin-spin interactions up to 2PN order. In addition, for the \texttt{SEOBNRv4T} model, the spin-induced quadrupole moment~\cite{Poisson1998} is also included. This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin.} \red{\sout{The model, however, does not account for the spin-induced quadrupole moment~\cite{Poisson1998}. This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, but is only significant for systems with large spin. The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{HarryHinderer2018}}}.

@@

@@

After rescaling with the total mass $M$, the aligned-spin EOB waveform with dynamical tides depends on 13 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless spin components along the orbital angular momentum $\chi_{A,B}$, the two spin-induced quadrupole moments $\bar Q_{A,B}$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$. To reduce the dimensionality of the intrinsic parameter space, we use nearly EOS-independent fits (universal relations) for these parameters in terms of $\Lambda_2$ as discussed below. This reduces the number of matter parameters from ten to two.

Eq.~(15) of Ref.~\cite{YagiYunes2017} provides a fit for $\bar Q$ as a function of $\Lambda_2$. This relation has been fit for a sample of EOSs and for tidal parameters in the range $1 \leq \Lambda_2\leq10^4$. This formula, however, diverges at small values of $\Lambda_2$, so we replace it with a polynomial function that approaches the Kerr value of $\bar Q = 1$ at $\Lambda_2 = 0$. We require this extension to be continuous at $\Lambda_2 = 1$. The exact relation we use is
\begin{equation}
\bar Q = \left\{\begin{array}{ll}
1 + f \Lambda_2 + g \Lambda_2^2 + h \Lambda_2^3\, , & 0 \leq \Lambda_2 \le 1 \\
e^{a + b \xi + c \xi^2 + d \xi^3 + e \xi^4}\, , & \Lambda_2 > 1
\end{array}\right.,
\end{equation}
where $\xi = \ln\Lambda_2$. The coefficients $\{a, b, c, d, e\} = \{ 0.194, 0.09163, 0.04812, -4.283\times 10^{-3}, 1.245\times 10^{-4} \}$ are from Table I of Ref.~\cite{YagiYunes2017}, and $\{f, g, h\} = \{0.42769, -0.32434, 0.11074\}$.

@@

BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the University of Washington for useful discussions at the start of this work. The authors were supported by... \red{[[Get AEI funding info.]]}

>>>>

@@

\subsection{Spin-induced quadrupole-monopole terms}

\red{[check notations have been introduced]} For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins. It includes consistently spin-orbit interactions up to 3.5PN order and spin-spin interactions up to 2PN order. In addition, the \texttt{SEOBNRv4T} model accounts for the spin-induced quadrupole moment of neutron stars~\cite{Poisson1998}, an extended-body effect distinguishing neutron stars from black holes. This effect is quadratic in the spins, appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin~\cite{HarryHinderer2018}. Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only, see~\cite{}.}, at the moment the effect is only consistently included at the leading 2PN order. We leave for future work an extension to higher orders\footnote{\red{complete this} We note that TaylorF2 models do include all spin-square terms at 3PN, see App.~\ref{sec:taylorf2}. They are also present in the IMRPhenomPv2+NRTides model, but not in the SEOBNRv4ROM+NRTides model. TEOBResumS is also limited to the leading 2PN order.}.

In the compact binary system we consider, the two material bodies, e.g. two neutron stars, with masses and spins $m_{A}$ and $\bs{S}_{A}$ for $A=1,2$, have spin-induced mass quadrupole moments $Q_{A,\mr{SS}}^{ij}=-\kappa_A S_A^{\langle i}S_A^{j\rangle}/m_A$ \cite{Poisson1998}, with $\kappa_{A}$ the quadrupole-monopole parameter. The case of two black holes is recovered for $\kappa_1=\kappa_2=1$, while for a neutron star $\kappa$ can be larger, of order 10 for hard equations of state.

The SEOB model should be modified as follows in order to take this into account at leading order. The SEOB effective Hamiltonian is related to the real Hamiltonian by (7.2) of~\cite{Barausse:2009xi}. Its structure is given by (5.70) of~\cite{Barausse:2009xi} as
\be
H_\mr{eff}=H_S+\beta^ip_i+\alpha\sqrt{\mu^2+\gamma^{ij}p_ip_j+\mc Q_4(p)}+H^\mr{BBH}_\mr{extra} \,,
\ee
where
\be
H^\mr{BBH}_\mr{extra}=\frac{1}{2r^3}(3n_in_j-\delta_{ij})\frac{\mu}{M}S_*^iS_*^j \,,
\ee
in terms of the tota mass $M$, the reduced mass $\mu = M \eta$, and the spin combination
\be	
	S_*^i=\frac{m_2}{m_1} S_1^i+\frac{m_1}{m_2}S_2^i \,.
\ee
We refer to~\cite{Barausse:2009xi} for the notations and meaning of the other terms. We find that, for systems including neutron stars, the latter contribution should be replaced with
\begin{align}
	H^\mr{BNS}_\mr{extra} &= \frac{1}{2r^3}(3n_in_j-\delta_{ij}) \nnm\\
	& \; \cdot \bigg[\frac{\mu}{M}S_*^iS_*^j+(\kappa_1-1)\frac{m_2}{m_1}S_1^i S_1^j+(\kappa_2-1)\frac{m_1}{m_2}S_2^i S_2^j\bigg] \label{eq:HBNSextra}
%\\\nnm
%&=\frac{1}{2r^3}(3n_in_j-\delta_{ij})\bigg[2\frac{\mu}{M}S_1^iS_2^j+\left(\kappa_1-1+\frac{m_2^2}{M^2}\right)\frac{m_2}{m_1}S_1^i S_1^j+\left(\kappa_2-1+\frac{m_1^2}{M^2}\right)\frac{m_1}{m_2}S_2^i S_2^j\bigg],
\end{align}
where we chose to use the spin variables $\bs S_1$ and $\bs S_2$ instead of $\bs S_*$ and $\bs S=\bs S_1+\bs S_2$ to avoid the occurence of the mass difference $m_{1} - m_{2}$ in denominators. This expression generalizes the black hole case and reduces to it for $\kappa_{1} = \kappa_{2} = 1$.

In the waveform, at the leading 2PN order, the only contribution to consider will be in the mode $h_{22}$. The total leading-order spin-squared contribution, for aligned-spin circular orbits, is given by~\cite{Kidder:1995zr, Will:1996zj, Buonanno:2012rv}
\be
h_{22}^\tr{LO-SS}=-\frac{8\nu M^3\omega^2}{R}\sqrt{\frac{\pi}{5}}e^{-2i\Phi}\mc A \,,
\ee
with $\Phi$ the orbital phase, $\omega = \dot{\Phi}$ the orbital frequency and $R$ the distance to the observer. Here, for a BBH,
\be\label{eq:ABBH}
\mc A^\mr{BBH}=\frac{1}{M^2}\left(\frac{S_1^2}{m_1^2}+2\frac{S_1S_2}{m_1m_2}+\frac{S_2^2}{m_2^2}\right) \,,
\ee
while for a BNS, this should be replaced with
\be\label{eq:ABNS}
\mc A^\mr{BNS}=\frac{1}{M^2}\left(\kappa_1\frac{S_1^2}{m_1^2}+2\frac{S_1S_2}{m_1m_2}+\kappa_2\frac{S_2^2}{m_2^2}\right) \,.
\ee
The total spin-squared contribution to the flux also includes squares of spin-orbit terms, and reads~\cite{Kidder:1995zr, Will:1996zj, Poisson1998}
\be
\mc F^\tr{LO-SS}=\frac{32\eta^2}{5} v^{14} \left[2\mc A+\frac{1}{16M^2}\left(\frac{S_1}{m_1}-\frac{S_2}{m_2}\right)^2\,\right] \,,
\ee
with $v=(M\omega)^{1/3}$.

When translating the above for the EOB factorized waveform (see e.g. (17) in~\cite{Taracchini:2012ig}), spin contributions in the effective source, tail factors and phase contribution would enter only at higher order, so that the only modification to consider is in the $\rho_{\ell m}$ factor for $\ell = 2$, $m = 2$. We have simply
\be\label{eq:rho22}
	\rho_{22}^{\textrm{LO-SS}} = \frac{1}{2}\mathcal{A} v^{4} \,,
\ee
with $\mathcal{A}$ the quantity given by~\eqref{eq:ABBH} for a BBH or~\eqref{eq:ABNS} for a BNS.

@@

After rescaling with the total mass $M$, the aligned-spin EOB waveform with dynamical tides depends on 13 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless spin components along the orbital angular momentum $\chi_{A,B}$, the two spin-induced quadrupole-monopole parameters $\kappa_{A,B}$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$. To reduce the dimensionality of the intrinsic parameter space, we use nearly EOS-independent fits (universal relations) for these parameters in terms of $\Lambda_2$ as discussed below. This reduces the number of matter parameters from ten to two.

Eq.~(15) of Ref.~\cite{YagiYunes2017} provides a fit for $\kappa$ as a function of $\Lambda_2$. This relation has been fit for a sample of EOSs and for tidal parameters in the range $1 \leq \Lambda_2\leq10^4$. This formula, however, diverges at small values of $\Lambda_2$, so we replace it with a polynomial function that approaches the Kerr value of $\kappa = 1$ at $\Lambda_2 = 0$. We require this extension to be continuous at $\Lambda_2 = 1$. The exact relation we use is
\begin{equation}
\kappa = \left\{\begin{array}{ll}
1 + f \Lambda_2 + g \Lambda_2^2 + h \Lambda_2^3\, , & 0 \leq \Lambda_2 \le 1 \\
e^{a + b \xi + c \xi^2 + d \xi^3 + e \xi^4}\, , & \Lambda_2 > 1
\end{array}\right.,
\end{equation}
where $\xi = \ln\Lambda_2$. The coefficients $\{a, b, c, d, e\} = \{ 0.194, 0.09163, 0.04812, -4.283\times 10^{-3}, 1.245\times 10^{-4} \}$ are from Table I of Ref.~\cite{YagiYunes2017}, and $\{f, g, h\} = \{0.42769, -0.32434, 0.11074\}$.

@@

BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the University of Washington for useful discussions at the start of this work. The authors were supported by... \red{[[Get AEI funding info.]]} \red{[Tanja Hinderer and Justin Vines derived the quadrupole-monopole terms for SEOB]}
